BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-16 Msc-Slides/main.tex
Slides re-worked

\section{Introduction}

@@

\begin{frame}
    \frametitle{The threat posed by evasive malware}

@@

\subsection{Function Call Monitoring}
\begin{frame}
    \frametitle{Function call monitoring}

Functions can abstract
implementation details providing
a semantically richer representation of some functionality. \\\bigskip
\textbf{Example}:
    \begin{figure}
    	\vspace{-0.4cm}
        \includegraphics[scale=0.6]{image/fcmonitoring}
    \end{figure}

\begin{beamerboxesrounded}[shadow=true]{}
The abstractions embodied by \textbf{system calls} and \textbf{library calls} \\ can be used to grasp the visible behavior of a malicious sample
\end{beamerboxesrounded}

\begin{frame}
    \frametitle{Implementation of function call monitoring}

	    \begin{beamerboxesrounded}[shadow=true]{API Hooking}
    The interception of function calls provided by dynamically
linked \\ libraries (DLLs)
    \end{beamerboxesrounded}
    \bigskip
	Three broad categories:

    \begin{itemize}
    \item Binary Rewriting
    \begin{itemize}
    \item Call Redirection
    \item Function Rewriting
    \end{itemize}
    \item Virtual Machine Introspection (VMI)
    \item \textbf{Dynamic Binary Instrumentation (DBI)}
    \end{itemize}

    \begin{textblock*}{2cm}(8.3cm,-3cm)
   \includegraphics[width=2cm]{image/hook.png}% use the \includegraphics command here
	\end{textblock*}

\begin{frame}[fragile]
    \frametitle{Dynamic Binary Instrumentation (DBI)}

@@

\subsection{Overview}
\begin{frame}
    \frametitle{Our solution: BlueTracer}

@@

\textcolor{sapienza}{\textbf{Building blocks:}}
\begin{itemize}
\item Based on the \textbf{Intel Pin} DBI framework
\item Integrated with the \textbf{BluePill} stealthy execution framework
%\item Combines reliable external sources of prototypes information
\end{itemize}

\bigskip
	    \begin{figure}
    	\vspace{-0.3cm}
    	%\hspace*{1cm}
        \includegraphics[width=7cm]{image/BluePill.pdf}
    \end{figure}

@@

\begin{frame}
    \frametitle{BlueTracer: challenges and features}
\setbeamerfont*{itemize/enumerate subbody}{parent=itemize/enumerate body}

\textcolor{sapienza}{\textbf{Implementation challenges:}}

\begin{itemize}
\item Heterogeneity of Windows libraries used in malware and the lack of well-structured documentation for their prototypes
\begin{itemize}
\item[$\rightarrow$] Integration of reliable external sources \\ (\textbf{Dr. Memory}
and \textbf{CISCO PyREBox})
\end{itemize}
\item Making best use of
Intel Pin's capabilities with respect to run-time CPU and memory costs
\item Addressing some inherent limitations of Intel Pin

@@

\textcolor{sapienza}{\textbf{Key features:}}
\begin{itemize}
\item Undetected tracing of input parameters, output buffers and return values of over 17 000 system calls and library calls
\item Logging of asynchronous events
\item Resolution of named constants
\end{itemize}

@@

\begin{frame}[fragile]
    \frametitle{Example of tracked evasion check}

\begin{beamerboxesrounded}[shadow=true]{}
Tracing a particular action of a malware instance allows to understand in detail what the sample's intentions are
\end{beamerboxesrounded}
\bigskip    
\textbf{Example}: file system artifacts can be checked in order to uncover the presence of a virtualized environment.
\vspace*{0.3cm}

	\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    %keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\linespread{1.3}\footnotesize\ttfamily,
    basewidth = {.48em},
   escapeinside={<@}{@>}
}
\begin{lstlisting}
~~3160~~ 24980 KERNELBASE.dll!GetFileAttributesW
24980 	arg 0: <@\alert{C:Windows\textbackslash system32\textbackslash drivers\textbackslash VBoxMouse.sys}@>
 (name=lpFileName, type=wchar_t*, size=0x2)
24980    executed KERNELBASE.dll!GetFileAttributesW =>
24980 	retval: 0xffffffff (name=Return value, type=DWORD, size=0x4)
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

    \begin{table}[h]
\vspace*{-0.1cm}
\begin{center}
\resizebox{9.2cm}{!}{
 \begin{tabular}{||c c c||}
 \hline
 \textbf{ID} & \textbf{MD5} & \textbf{Name} \\ 
 \hline\hline
 1 & 0af4ef5069f47a371a0caf22ae2006a6 & \textit{trojan/banker} \\ 
 \hline
 2 & 9437eabf2fe5d32101e3fbf9f6027880 & \textit{dropper} \\
 \hline
 3 & cbdda646a20d95f078393506ecdc0796 & \textit{trojan}\\
 \hline
 4 & cfdd16225e67471f5ef54cab9b3a5558 & Olympic\\
 \hline
 5 & ef694b89ad7addb9a16bb6f26f1efaf7 & CCleaner\\ 
 \hline
\end{tabular}
}
\end{center}
\end{table}
\vspace{-0.1cm}
Evaluation was done manually and is a time-consuming process:
\begin{itemize}
\item Check if logs are congruous with Joe Security reports
\item Process Monitor as ground truth for system activity
\end{itemize}
\medskip
\begin{beamerboxesrounded}[shadow=true]{}
The
logs collected by BlueTracer reveal behaviors consistent with the analysis reports
authored by Joe Security
\end{beamerboxesrounded}

@@

	\textbf{\textcolor{sapienza}{Contribution:}} \\
	\smallskip	
	\tab Design and implementation of \textbf{BlueTracer},
a robust library and system call tracer for Windows programs specialized in evasive
malware.
	\\\bigskip
	\textbf{\textcolor{sapienza}{Future Developments:}} \\
	\begin{itemize}
	\item Adoption of an automatic methodology for large-scale evaluation 
	\item Improvement of logging capabilities
	\item Usage of log filtering techniques
	\end{itemize}

>>>>

\section{Intro.}

@@

\subsection{Function Call Monitoring}
\begin{frame}[fragile]
    \frametitle{Function call monitoring}

\begin{itemize}
\item Functions can abstract
implementation details providing
a semantically richer representation of some functionality
\item The abstractions embodied by \textbf{system calls} and \textbf{library calls} \\ can be used to grasp the visible behavior of a malicious sample 
\end{itemize}
\medskip
\textcolor{sapienza}{\textbf{Example}}:\\
\bigskip

\begin{beamerboxesrounded}[shadow=true]{}

\lstset{
    language=C++,
    stringstyle=\color{sapienza}, % string color
    basicstyle=\linespread{1.5}\ttfamily\small,
}

\begin{lstlisting}[xleftmargin=0pt]
RegCreateKey("...\CurrentVersion\Run\monitor") 
CreateDirectory("C:\Windows\utils")
CreateFile("C\Windows\utils\GFypmMVqJQOEQqy.exe")
\end{lstlisting}

\end{beamerboxesrounded}

\begin{frame}[fragile]
    \frametitle{Implementation of function call monitoring}

	    \begin{beamerboxesrounded}[shadow=true]{API Hooking}
    The interception of function calls provided by dynamically
linked \\ libraries (DLLs)
    \end{beamerboxesrounded}
    \bigskip

%	Three broad categories:

%    \begin{itemize}
%    \item Binary Rewriting
%    \begin{itemize}
%    \item Call Redirection
%    \item Function Rewriting
%    \end{itemize}
%    \item Virtual Machine Introspection (VMI)
%    \item \textbf{Dynamic Binary Instrumentation (DBI)}
%    \end{itemize}

%    \begin{textblock*}{2cm}(8.3cm,-3cm)
%   \includegraphics[width=2cm]{image/hook.png}% use the \includegraphics command here
%	\end{textblock*}

One technique is \textbf{Dynamic Binary Instrumentation (DBI)}.
\begin{itemize}
\item[$\rightarrow$] The behaviour of an application is inspected at run-time, without the need of recompiling it, via the injection of analysis code
\end{itemize}
\smallskip

\begin{block}{}
\begin{lstlisting}[basicstyle=\ttfamily\large,xleftmargin=50pt]
<@\textcolor{red}{record\_before(libcall\_name, arg1)}@> 
retval = libcall(arg1, &arg2)
<@\textcolor{red}{record\_after(retval, *arg2)}@> 
\end{lstlisting}
\end{block}

\iffalse
\begin{frame}[fragile]
    \frametitle{Dynamic Binary Instrumentation (DBI)}

\subsection{Evasion}
\begin{frame}
    \frametitle{The threat posed by evasive malware}

@@

\begin{frame}
    \frametitle{Existing problems}

\section{Problem}	
Two main problems which need to be addressed:
\\\bigskip
\begin{enumerate}
\item Existing API tracing tools have \textbf{limited logging capabilities}
\smallskip
\begin{itemize}
\item[--] The amount of recorded information is too little
\item[--] It is hard to distinguish the calls made directly from the sample's main executable from the ones made by libraries
\end{itemize}
\bigskip
\item Current API hooking techniques are \textbf{easily detectable} and are not coupled with mechanisms to hide their presence
from evasive malware
\end{enumerate}

@@

\subsection{Overview}
\begin{frame}
    \frametitle{BlueTracer}
    \setbeamerfont*{itemize/enumerate subbody}{parent=itemize/enumerate body}

@@

\textcolor{sapienza}{\textbf{The tool possesses a remarkable logging power:}}
\begin{itemize}
\item Access to a source of calls related information is required
\item Logging operations are made challenging by the heterogeneity of Windows libraries used in malware and the lack of well-structured documentation for their prototypes
\begin{itemize}
\item[$\rightarrow$] Integration of reliable external sources \\ (\textbf{Dr. Memory}
and \textbf{CISCO PyREBox})
\end{itemize} 
\end{itemize}

\textcolor{sapienza}{\textbf{Key features:}}
\begin{itemize}
\item Undetected tracing of input parameters, output buffers and return values of over 17 000 system calls and library calls
\item Logging of asynchronous events
\item Allows to trace only calls from the main executable
\end{itemize}

@@

\begin{frame}
    \frametitle{BlueTracer (cont'd)}
\setbeamerfont*{itemize/enumerate subbody}{parent=itemize/enumerate body}

\textcolor{sapienza}{\textbf{Solution to the detection problem:}}
\begin{itemize}
\item Seamless integration with \textbf{BluePill}, a stealthy execution framework based on \textbf{Intel Pin}
%\item Combines reliable external sources of prototypes information
\end{itemize}

%\textcolor{sapienza}{\textbf{Building blocks:}}
%\begin{itemize}
%\item Based on the \textbf{Intel Pin} DBI framework
%\item Integrated with the \textbf{BluePill} stealthy execution framework
%\item Combines reliable external sources of prototypes information
%\end{itemize}

\bigskip
	    \begin{figure}
    	\vspace{-0.3cm}
    	%\hspace*{1cm}
        \includegraphics[width=6cm]{image/BluePill.pdf}
    \end{figure}

\textcolor{sapienza}{\textbf{Faced implementation challenges:}}
\begin{itemize}
\item Making best use of
Intel Pin's capabilities with respect to run-time CPU and memory costs
\item Addressing some inherent limitations of Intel Pin

@@

@@

@@

    \begin{table}[h]
\vspace*{-0.1cm}
\begin{center}
\resizebox{9.2cm}{!}{
 \begin{tabular}{||c c c||}
 \hline
 \textbf{ID} & \textbf{MD5} & \textbf{Name} \\ 
 \hline\hline
 1 & 0af4ef5069f47a371a0caf22ae2006a6 & \textit{banker} \\ 
 \hline
 2 & 9437eabf2fe5d32101e3fbf9f6027880 & \textit{dropper} \\
 \hline
 3 & cbdda646a20d95f078393506ecdc0796 & \textit{trojan}\\
 \hline
 4 & cfdd16225e67471f5ef54cab9b3a5558 & Olympic\\
 \hline
 5 & ef694b89ad7addb9a16bb6f26f1efaf7 & CCleaner\\ 
 \hline
\end{tabular}
}
\end{center}
\end{table}
\vspace{-0.1cm}
Evaluation was done manually and is a time-consuming process:
\begin{itemize}
\item Check if logs are congruous with Joe Security reports
\item Process Monitor as ground truth for system activity
\end{itemize}
\medskip
\begin{beamerboxesrounded}[shadow=true]{}
The
logs collected by BlueTracer reveal behaviors consistent with the analysis reports
authored by Joe Security
\end{beamerboxesrounded}

\begin{frame}[fragile]
    \frametitle{Example of tracked malevolent action}

\begin{beamerboxesrounded}[shadow=true]{}
Tracing a particular action of a malware instance allows to understand in detail what the sample's intentions are
\end{beamerboxesrounded}
\bigskip    
\textbf{Example}: dropping a malicious executable
\vspace*{0.3cm}

	\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    %keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\linespread{1.3}\footnotesize\ttfamily,
    basewidth = {.48em},
   escapeinside={<@}{@>}
}

\begin{lstlisting}
~~1116~~ 138 kernel32.dll!CopyFileA
138 	arg 0: c:\Users\Simuset\Desktop\sample1.exe 
(name=lpExistingFileName, type=char*, size=0x1)
138 	arg 1: <@\alert{C:\textbackslash Windows\textbackslash system32\textbackslash $\dagger$\textbackslash ffpb6966.exe}@> 
(name=lpNewFileName, type=char*, size=0x1)
138 	arg 2: 0x0 (name=bFailIfExists, type=(long/int), size=0x4)
138    executed kernel32.dll!CopyFileA =>
138 	retval: 0x1 (name=Return value, type=(long/int), size=0x4)
\end{lstlisting}

\iffalse
\begin{lstlisting}
~~972~~ 1397 kernel32.dll!CreateFileW
1397 	arg 0: <@\alert{C:\textbackslash Windows  GFypmMVqJQOEQqy.exe}@> 
(name=lpFileName, type=wchar_t*, size=0x2)
1397 	arg 1: 0x40000000 (name=dwDesiredAccess, type=DWORD, size=0x4)
1397 	arg 2: 0x1 (name=dwShareMode, type=DWORD, size=0x4)
1397 	arg 3: 0x00000000 (name=lpSecurityAttributes, type=struct*|_SECURITY_ATTRIBUTES, size=0x60)
1397 	arg 4: 0x2 
(name=dwCreationDisposition, type=DWORD, size=0x4)
1397 	arg 5: 0x0 (name=dwFlagsAndAttributes, type=DWORD, size=0x4)
1397 	arg 6: 0x0 (name=hTemplateFile, type=DWORD, size=0x4)
1397    executed kernel32.dll!CreateFileW =>
1397 	retval: 0x128 (name=Return value, type=DWORD, size=0x4))
\end{lstlisting}
\fi

\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

	\textbf{\textcolor{sapienza}{Contribution:}} \\
	\smallskip	
	\tab Design and implementation of \textbf{BlueTracer},
a robust library and system call tracer for Windows programs specialized in evasive
malware.
	\\\bigskip
	\textbf{\textcolor{sapienza}{Future Developments:}} \\
	\begin{itemize}
	\item Automatic methodology for large-scale evaluation 
	\item Improvement of logging capabilities
	\item Usage of log filtering techniques
	\end{itemize}
